Tail Flexibility in the Degrees of Preferential

### 2025-05-19 paper.qmd
Restructuring paper

The degree distribution of a network can be very informative about the networks structure, giving some understanding about the inequality in the influence of the nodes. Often the degree distribution is one of the simplest places to look when trying to understand this, as the full evolution of the network is usually not available. This makes modelling the degree distribution an important topic in research with contributions such as [ref].

When looking at the degree distributions of real networks, a power law is an attractive option to use when modelling as it is seemingly observed in a lot of networks. Fitting the discrete power law to the degree distribution come with the added benefit of suggesting preferential attachment as the generating mechanism since it has been shown to generate networks with a power law degree distribution.

If the power law is an adequate model to use for real degree distributions has become a heated debate, with alternatives such as @Broido_2019 seeming to outperform the power law in some cases. Additionally, @lef24 find that for many real degree distributions the body may follow a power law, the right tail does not or at least seems to be lighter than is implied by the body whilst still being heavy. One method for modelling the tail of the degree distributions of real networks is to use a discretised variation of the generalised Pareto distribution (GPD) usually dubbed the Integer GPD (GPD) where $X|X>v \sim \mathrm{IGPD}(\xi,\sigma,v)$ is defined by the survival:

@@

[CL: Is the precise form of the IGPD needed here? The big picture is probably more important: extreme value methods have not been used in analysing degree distributions until recently, such as @vvvk19 and @lef24. Not using these methods essentially downweights the deviation of the vertices with large degrees, which have far more influence on the network than those with small degrees, from the power law behaviour.]

@shimura12 introduces a quantity that will help in determining what domain of attraction a discrete distribution belongs to, that is:

@shimura12 then states that if $\lim_{n\rightarrow\infty} \Omega(F,n) = 1/\alpha$ ($\alpha>0$), then $F$ is heavy tailed with $\bar F(n) \sim n^{-\alpha}$. Additionally, if $\lim_{n\rightarrow\infty} \Omega(F,n) = 0$ then the distribution is light tailed.

[CL: The results of @shimura12 could be moved closer to Proposition 2.1. In the introduction, if you want to include @shimura12, you can mention the paper in passing, when outlining your approach, e.g. "Using the results by @shimura12, we derive the tail index / heaviness of the limiting degree distribution of the GPA model [@rudas07]. This informs us the precise class of preference functions that result in flexible heavy tail. "]

These models that seem to outperform the power law do not however come with the added benefit of suggesting a generating mechanism for the networks in the same way the power law does. Efforts to make the preferential attachment model more flexible have not yet remedied this issue with @krapivsky01 finding that using a sub/super-linear preference function yields a Weibull tail and a degenerate network, where one vertex eventually gains all new edges, respectively.

The network generative model that we will be focussing on in this paper is dubbed General Preferential Attachment in @rudas07 and is defined as follows:

Starting at time $t=0$ with an initial network of $m$ vertices that each have no edges, at times $t=1,2,\ldots$ a new vertex is added to the network bringing with it $m$ directed edges (with the new vertex as the source); the target for each of these edges are selected from the vertices already in the network with weights proportional to some function $b$ of their degree.

Estimating the preference function used by network generative models like this is of great interest, with the potential to provide a deeper understanding of the relationships between the vertices in the network. Several papers and at least one R package (`PAfit`) have been dedicated to estimating the preference function that may have been used when a network was growing; one thing all these have in common is requiring access to a networks evolution data, something that is not always readily available or feasible to access. A method for estimating a preference function without this data available would be extremely useful.

[@rudas07] introduces theoretical results for this model with a general preference function in the cases that the network being generated is a tree ($m=1$); presenting an opportunity to design a preference function that will capture the behaviours observed in the degree distributions of real networks. Since they provide theoretical results for the limiting degree distribution, a model of this form would be able to be fitted to the degree distributions of real networks where the full evolution is not available, only a snapshot.

@@

# General Preferential Attachment {#sec-gpa}

The focus of this section is the limiting degree distribution of a preferential attachment network with preference function $b(\cdot)$ where each vertex brings with it $m$ edges. This preference function is subject to the following conditions:

$$
b:\mathbb N \mapsto \mathbb R^+\setminus\{0\},
$$

@@

Given these two conditions an expression for the survival of the limiting degree distribution can be found in the case that $m=1$; obtained by considering a branching process that is equivalent to the growth of the network, as in [@rudas07]. Theorem 1 from [@rudas07] states that for the tree $\Upsilon(t)$ at time $t$:

@@

$$
f(k) = \frac{\lambda^*}{\lambda^* + b(k)}\prod_{i=0}^{k-1}\frac{b(i)}{\lambda^*+b(i)},\qquad k\in\mathbb N.
$$

which aligns with the result from [@rudas07], where the expression is derived using the same branching process results.

Using results from [@shimura12] the domain of attraction of this limiting degree distribution can be found.

@@

$$
\lim_{k\rightarrow\infty}\Omega(F,n) = \lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{\lambda^*}.
$$
:::

@prp-omega implies that the only way to obtain a heavy tailed degree distribution using a non-decreasing preference function is to have that preference function be asymptotically linear. The next subsection focuses on a particular class of these preference functions that provide flexible behaviour but guarantee a  heavy tail.

## Preferential Attachment with flexible heavy tail {#sec-model}

Previously, in [@krapivsky01], a preference function of the form $k^\alpha +\varepsilon$ has been considered. However, it has been noted that super-linear preferential attachment ($\alpha > 1$) leads to a degenerate limiting degree distribution where at some point one vertex gains the connection from every vertex that joins the network, additionally sub-linear preferential attachment ($\alpha <1$) has been shown to lead to a light tailed limiting degree distribution. Consider a preference function of the form:

@@

@@

The survival (@eq-polysurv) can be connected to the IGPD mentioned in [@sec-intro] by using Sterling's approximation of the gamma function to obtain:

<!-- $$ -->
<!-- \frac{\Gamma(x+y)}{\Gamma(x)}\approx x^{y}, -->
<!-- $$ -->
<!-- and utilising the expression from @eq-polysurv: -->

<!-- \begin{align*} -->
<!-- \bar F(k|k\ge k_0) &= \frac{\Gamma\left(\frac{\lambda^* + k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(\frac{k_0^\alpha + \varepsilon}{\beta}\right)}\times\frac{\Gamma\left(k-k_0  +1 + \frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0  +1 + \frac{\lambda^*+ k_0^\alpha + \varepsilon}{\beta}\right)}\\ -->
<!-- &\approx\left(\frac{k_0^\alpha+\varepsilon}{\beta}\right)^{\lambda^*/\beta}\left(k-k_0+1+\frac{k_0^\alpha + \varepsilon}{\beta}\right)^{-\lambda^*/\beta}\\ -->
<!-- &=\left(\frac{k_0^\alpha+\varepsilon}{k_0^\alpha+\varepsilon + \beta}\right)^{\lambda^*/\beta}\left(\frac{\beta(k-k_0)}{\beta + k_0^\alpha+\varepsilon} + 1\right)^{-\lambda^*/\beta}\\ -->
<!-- &=\left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^{*}/\beta} -->
<!-- \end{align*} -->

@@

### Piecewise Linear Model

When $\alpha = 1$ the preference function simplifies to a piecewise linear function with $\beta$ representing the ratio of the gradients of the two components:

$$
b(k) = \min(k,k_0) + \varepsilon + \mathbb I\{k\ge k_0\}\beta(k-k_0)
$$ for some $\beta, \varepsilon, k_0>0$.

Following @eq-polysurv the limiting degree distribution is:

$$
\bar F(k) = \begin{cases}
\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k+1+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k+1+\lambda^*+\epsilon)},&k<k_0\\
\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k_0+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k_0+\lambda^*+\varepsilon)}\times \frac{\Gamma\left(\frac{k_0+\varepsilon+\lambda^*}{\beta}\right)}{\Gamma\left(\frac{k_0+\varepsilon}{\beta}\right)}\times\frac{\Gamma\left(k-k_0 + \frac{k_0+\varepsilon}{\beta}+1\right)}{\Gamma\left(k-k_0 + \frac{k_0+\varepsilon+\lambda^*}{\beta}+1\right)},&k\ge k_0
\end{cases}
$$ {#eq-linsurv}

Just like the previous model, the survival can be approximated by:

$$
\bar F(k) \approx \begin{cases}
\frac{\Gamma(\lambda^*+\varepsilon)}{\Gamma(\varepsilon)} (1+\varepsilon)^{\lambda^*}\left(\frac{k}{1+\varepsilon} + 1\right)^{-\lambda^*},&k<k_0\\
\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k_0+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k_0+\lambda^*+\varepsilon)}\times\left(\frac{\beta(k+1-k_0)}{k_0+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0
\end{cases}
$$

Some examples of the possible degree distributions are shown in the central row of @fig-polylinsurv.

@@

surfit_list = list()
pars = readRDS('results/recovery_pars.rds')
recovery_list = readRDS('results/recovery_dat.rds')
selected = c(1,27,14,35)
# selected = 1:36
for(k in 1:length(selected)){
  j=selected[k]
  x = recovery_list[[j]]$mcmc$dat[,1]
  ls = c()

@@

```{r fig-rec2, out.width="80%"}
#| echo: false
#| fig.cap: Posterior estimates of parameters for data simulated from the proposed model with various combinations of ($\alpha$,$\beta$,$\varepsilon$) and $k_0=20$.
knitr::include_graphics("mcmc_plot.jpg")
```

@@

Turning now to real data, the goal is to fit the model to the degree distributions of real networks from various sources. [describe networks being used]. Alongside fitting this model to the degree distributions, we then compare the fit to that of an mixture distribution that was used in @lef24 and is similar to others that have been used to model degree distributions [others].

@@

plt1 = readRDS('results/comp_plot.rds')

@@

plt2 = readRDS('results/shape_plot.rds')

@@

plt3 = readRDS('results/PA_plot.rds')
plot(plt3)

@@

```{r fig-rec3, out.width="80%"}
#| fig-cap: Posterior estimates of parameters for real data.
#| echo: false
knitr::include_graphics("pars_plot.png")
```

@@

>>>>

Networks appear in many fields such as sociology, politics, epidemiology, and economics. Statistical methods have been and continue to be used to study networks that appear in these fields and beyond, from the use of stochastic block models to detect communities within the French political blogosphere [@Latouche11], to using Exponential Random Graph Models (ERGMs) to analyse the global trade network [@Setayesh22], and the use of mechanistic models in neuroscience to investigate the patterns of connections in neural systems [@Betzel17]. One such mechanistic model is the general preferential attachment (PA) model where as a network grows and new vertices join they connect to those already present in the network with weights proportional to some preference function $b$ of their degree. The well-known Barabási-Albert (BA) model is a special case of this when $b\sim k$ [BA99].

Mechanistic models, like the PA model, are in general only particularly useful when simulating networks and not fitting to the real data as usually only snapshots of real networks are available and not the entire evolution. As such, one of the most informative statistics when regarding a network is the degree distribution. When looking at the degree distributions of real networks it is attractive to many to use the power law to model the degree distribution, as the BA model generates degrees that follow a power law, and therefore claim that since the degree distribution is roughly power law then the underlying growth mechanism may be preferential attachment such as the BA model.  However, it has been shown that a general preference function will not always lead to a power law degree distribution [@krapivsky01] and additionally it has been debated for decades whether most real data actually follows a power law [@Voitalov_2019] or not[@Broido_2019].

Throughout this debate the use of methods from extremes has been absent until relatively recently with the papers such as [@Voitalov_2019] and [Lee]. Forgoing the use of extreme value methods essentially downplays how deviations from the power law in the largest degrees may affect the network given that they are often much more influential than the vertices with smaller degrees.

Noting this, many have now started to use these extreme value methods when analysing networks. [@wan19] showed that extreme value estimation methods prove more robust than existing parametric approaches especially when the data are corrupted or the model has been misspecified, demonstrating that extreme value methods can be a valuable tool when studying networks. [@wang2022random] investigates a PA model with a affine preference function (similar to the BA model) with the added behaviour of heterogeneous reciprocal edges, they show that the distribution of in-degrees and out-degrees produced by such a model is multivariate regularly varying and they also have hidden regular variation. [Lee] using methods from discrete extremes modelled the right tail of network degrees with a  discretised variation of the generalised Pareto distribution (GPD) usually dubbed the Integer GPD (IGPD) where $X|X>v \sim \mathrm{IGPD}(\xi,\sigma,v)$ is defined by the survival:

@@

The study of degree distributions, whether using extremes or not, usually does not reveal any information about the preference function even if PA was the underlying generating mechanism of the network. Here we address this gap in the literature, more specifically investigating if given the degree distribution of a network assumed to come from the PA model can we directly infer the model parameters? 
We build upon the foundation provided by [@rudas07] deriving the limiting degree distribution in terms of the preference function and suggest a class of preference functions that result in realistic tails that are heavy but are lighter than implied by the power law.

@@

# PA model

The network generative model that we will be focussing on in this paper is dubbed General Preferential Attachment in [@rudas07] and is defined as follows:

Starting at time $t=0$ with an initial network of $m$ vertices that each have no edges, at times $t=1,2,\ldots$ a new vertex is added to the network bringing with it $m$ directed edges (with the new vertex as the source); the target for each of these edges are selected from the vertices already in the network with weights proportional to some function $b$ of their degree where the preference function $b$ is chosen such that:

$$
b:\mathbb N \mapsto \mathbb R^+\setminus\{0\},
$$ {#eq-condb1}

@@

Given conditions [@eq-condb1] and [@eq-condb2] an expression for the survival of the limiting degree distribution can be found in the case that $m=1$; obtained by considering a branching process that is equivalent to the growth of the network, as in [@rudas07]. Theorem 1 from [@rudas07] states that for the tree $\Upsilon(t)$ at time $t$:

@@

$$
f(k) = \frac{\lambda^*}{\lambda^* + b(k)}\prod_{i=0}^{k-1}\frac{b(i)}{\lambda^*+b(i)},\qquad k\in\mathbb N.
$$
We are specifically interested in how the tail heaviness ($\xi$ in the IGPD) is affected by the preference function $b$, [@shimura12] introduces a quantity that will help in determining what domain of attraction this discrete distribution belongs to, that is:

[@shimura12] then states that if $\lim_{n\rightarrow\infty} \Omega(F,n) = 1/\alpha$ ($\alpha>0$), then $F$ is heavy tailed with $\bar F(n) \sim n^{-\alpha}$. Additionally, if $\lim_{n\rightarrow\infty} \Omega(F,n) = 0$ then the distribution is light tailed. This allows us to show the following

@@

$$
\lim_{k\rightarrow\infty}\Omega(F,n) = \lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{\lambda^*}.
$$
:::
See [appendix] for the details of the proof.

@prp-omega aligns with the result from [@krapivsky01] demonstrating that a sub-linear preference function will lead to a light tailed distribution. So, in order for the degree distribution to be heavy tailed we need that the limit $\lim_{k\rightarrow\infty} b(k+1)-b(k)$ exists and is positive. We can in fact show that BA model produces a heavy tailed degree distribution with index $\xi=0.5$ by considering the preference function $b(k) = k + \varepsilon$,  noting that the limit $\lim_{k\rightarrow\infty} b(k+1)-b(k)$ is simply $\lim_{k\rightarrow\infty} 1 = 1$ leaving the tail heaviness to be $1/\lambda^*$ which using $\hat\rho$ can be found to be $1/2$.

We can show that in order for the degree distribution to be heavy tailed, the preference function must be asymptotically linear i.e. $\lim_{k\rightarrow\infty}\frac{b(k)}{k} =  c>0$.

Consider the definition of heavy tails:

\begin{align*}
\lim_{k\rightarrow\infty}\Omega(F,n) &= c>0\\
\lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{\lambda^*} &= c\\
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] &= \lambda^*c\\
\lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{(k+1)-(k)} &= \lambda^*c\\
\end{align*}

by Stolz-Cesaro Theorem:

$$
\lim_{k\rightarrow\infty}\frac{b(k)}{k} = \lambda^*c > 0.
$$

Since $b$ must be asymptotically linear in order to have positive tail heaviness we propose the following preference function:

@@

This preference function, as per @prp-omega, will produce a degree distribution with tail heaviness $\beta/\lambda^*$ guaranteeing a heavy tail. We study this preference function further in the next section.

# Preferential Attachment with flexible heavy tail {#sec-model}

@@

The survival (@eq-polysurv) can be connected to the IGPD mentioned in [@sec-intro] by using Stirling's approximation of the gamma function to obtain:

$$
\frac{\Gamma(x+y)}{\Gamma(x)}\approx x^{y},
$$
and utilising the expression from @eq-polysurv:

\begin{align*}
\bar F(k|k\ge k_0) &= \frac{\Gamma\left(\frac{\lambda^* + k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(\frac{k_0^\alpha + \varepsilon}{\beta}\right)}\times\frac{\Gamma\left(k-k_0  +1 + \frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0  +1 + \frac{\lambda^*+ k_0^\alpha + \varepsilon}{\beta}\right)}\\
&\approx\left(\frac{k_0^\alpha+\varepsilon}{\beta}\right)^{\lambda^*/\beta}\left(k-k_0+1+\frac{k_0^\alpha + \varepsilon}{\beta}\right)^{-\lambda^*/\beta}\\
&=\left(\frac{k_0^\alpha+\varepsilon}{k_0^\alpha+\varepsilon + \beta}\right)^{\lambda^*/\beta}\left(\frac{\beta(k-k_0)}{\beta + k_0^\alpha+\varepsilon} + 1\right)^{-\lambda^*/\beta}\\
&=\left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^{*}/\beta}
\end{align*}

@@

For a given network with degree count vector $\pmb n = (n_0, n_1, \ldots, n_M)$ and maximum degree $M$ the likelihood is given by:

\begin{align*}
L(\pmb x,\pmb n | \pmb \theta) = \left(\frac{\lambda^*}{\lambda^*+\varepsilon}\right)^{n_0}\left(\prod_{j=l}^{k_0-1}\frac{j^\alpha +\varepsilon}{\lambda^* + j^\alpha +\varepsilon}\right)^{\left(\sum_{i\ge k_0}n_{i}\right)} \prod_{l \le i<k_0}\left(\frac{\lambda^*}{\lambda^* +i^\alpha + \varepsilon } \prod_{j=l}^{k_0-1}\frac{j^\alpha + \varepsilon}{\lambda^* + j^\alpha + \varepsilon}\right)^{n_i}\\ \times \prod_{i\ge k_0}\left(\frac{\text{B}(i-k_0 + (k_0^\alpha + \varepsilon)/\beta,1+\lambda^*/\beta)}{\text{B}((k_0^\alpha + \varepsilon)/\beta,\lambda^*/\beta)}\right)^{n_i}
\end{align*}

where $B(y,z)$ is the the beta function and $l\ge0$ variable that allows truncating the data such that the minimum degree is $l$.

This likelihood allows for the parameters to be inferred using either a frequentist or a Bayesian approach.

@@

surfit_list = list()
pars = readRDS('recovery_pars.rds')
recovery_list = readRDS('recovery_dat.rds')
selected = c(1,27,14,35)
# selected = 1:36
for(k in 1:length(selected)){
  j=selected[k]
  x = recovery_list[[j]]$mcmc$dat[,1]
  ls = c()

@@

:::{#fig-rec2 out.width="80%"}
![: Posterior estimates of paramters for data simulated from the proposed model with various combinations of ($\alpha$,$\beta$,$\varepsilon$) and $k_0=20$.](mcmc_plot.jpg)
:::

@@

Turning now to real data, the goal is to fit the model to the degree distributions of real networks from various sources. [describe networks being used]. Alongside fitting this model to the degree distributions, we then compare the fit to that of an mixture distribution that was used in [clement] and is similar to others that have been used to model degree distributions [others].

@@

plt1 = readRDS('comp_plot.rds')

@@

plt2 = readRDS('shape_plot.rds')

@@

plt3 = readRDS('PA_plot.rds')
plot(plt3)

@@

:::{#fig-rec2 out.width="80%"}
![: Posterior estimates of paramters for real data.](pars_plot.png)
:::

@@

If this were the case, the posterior distributions of parameters from @sec-real could be interpreted as the approximate distributions from which a vertex's fitness values are drawn instead of providing estimates of the universal parameter values. Additionally, it allows the ability to set some parameters as universal and some as vertex-wise for example one could let $\alpha, \beta, k_0$ be the same for all vertices but let each vertex in the network have its own "zero appeal" $\varepsilon_i\overset{\text{iid}}{\sim} \text{Ga}(\theta_1, \theta_2)$; moving the model to being more realistic and accounting for the fact that in many real networks the objects vertices represent have some value instrinsic to themselves other than the number of connections they have.
